2.4 PRNT50

### 2024-08-08 2_scripts/2.4 PRNT50.qmd
Codes for vax%

tb_abbs<-APLs %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbs$prop.abbs1<-tb_abbs$pos_abbott_s/tb_abbs$n
tb_abbs$prop.abbs2<-tb_abbs$pos_abbott_s2/tb_abbs$n

@@

```{r}
# create vax% cat var
APLs$vax_cat<-rep(NA, 214780)
APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
                             "4 Doses")]<-1
APLs$vax_cat[APLs$vax %in% c("Not Immunized", "Pre-Immunization")]<-0
table(APLs$vax, APLs$vax_cat, useNA = 'ifany')
```

```{r}
# create merged data
tb_vax<-APLs %>%
  .[.$abb_s_AU==1,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_vax= sum(vax_cat, na.rm = T)
            )
tb_vax$prop.vax<-tb_vax$n_vax/tb_vax$n

@@

**Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.

@@

APLs$vax_est<-rep(0, 214780)
APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
APLs$vax_est[is.na(APLs$pos_abbott) | is.na(APLs$pos_abbott_s)]<-NA
table(APLs$vax_est)
```

```{r}
tb_vax_est<-APLs %>%
#  .[!is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_vax= sum(vax_est, na.rm = T)
            )
tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n

# clean NAs
tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]

@@

d01<-tb_vax_AB
colnames(d01)<-c("samplemonth", "prop.vax", "data")
```

@@

```{r}
# Estimated vaccination rate: by anti-S + and anti-N neg
View(tb_vax_est)
d02<-select(tb_vax_est, c(1,4))
d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))

# Estimated vaccination rate: by anti-S only
View(tb_abbs)
d03<-select(tb_abbs, samplemonth, prop.abbs1, study)
colnames(d03)<-c("samplemonth", "prop.vax", "data")
d03$data<-rep("S_pos", length(d03$samplemonth))

@@

# merge all
dta.vax<-rbind.data.frame(d01, d02, d03, d04)
rm(d01, d02, d03, d04)

dta.vax$prop.vax[dta.vax$data=="AB_general"]<-dta.vax$prop.vax[dta.vax$data=="AB_general"]/100
```

@@

```{r}
p12<-dta.vax %>%
  ggplot(aes(x=samplemonth, y=prop.vax, group=data, colour=data)) +
  geom_point(aes(colour=data)) +
  geom_line(aes(colour=data))  +
  theme_bw() + 
  ylim(c(0,1)) +
  ylab("Vaccinated%")
p12
```

>>>>

APLs$pos_abbott_s_na<-is.na(APLs$pos_abbott_s)
tb_abbs<-APLs %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_na_s = sum(pos_abbott_s_na, na.rm = T),
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbs$prop.abbs1<-tb_abbs$pos_abbott_s/(tb_abbs$n-tb_abbs$n_na_s)
tb_abbs$prop.abbs2<-tb_abbs$pos_abbott_s2/(tb_abbs$n-tb_abbs$n_na_s)
tb_abbs$prop.abbs1[1:9]<-rep("NA",9)
tb_abbs$prop.abbs2[1:9]<-rep("NA",9)

@@

```{r}
# create vax% cat var
APLs$vax_cat<-rep(NA, length(APLs$clean_IDe))
APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
                             "4 Doses")]<-1
APLs$vax_cat[APLs$vax %in% c("Not Immunized", "Pre-Immunization")]<-0
table(APLs$vax, APLs$vax_cat, useNA = 'ifany')
```

```{r}
# create merged data
tb_vax<-APLs %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_vax= sum(vax_cat, na.rm = T)
            )
tb_vax$prop.vax<-tb_vax$n_vax/tb_vax$n

@@

**Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted. An alternative solution is to estimate vax% using (S+ N-) by (N-), assuming vaccination is not related to anti N positivity (which is probably not true and we can cross-verify using our own data).

@@

APLs$vax_est<-rep(0, length(APLs$clean_IDe))
APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
APLs$vax_est[is.na(APLs$pos_abbott) | is.na(APLs$pos_abbott_s)]<-NA
table(APLs$vax_est)
```

```{r}
# Modify APLs data, add na indicator, N=69108
table(APLs$pos_abbott, useNA = "ifany")
APLs$pos_abbott_na <- is.na(APLs$pos_abbott)
table(APLs$pos_abbott, APLs$pos_abbott_na, useNA = 'ifany')

tb_vax_est<-APLs %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_pos = sum(pos_abbott, na.rm = T),
            n_na = sum(pos_abbott_na, na.rm=T),
            n_vax= sum(vax_est, na.rm = T)
            )
tb_vax_est$n_neg = tb_vax_est$n - tb_vax_est$n_pos - tb_vax_est$n_na
tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n_neg

# clean NAs
tb_vax_est$prop.vax[1:17]<-rep(NA, 17)
tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]

@@

d01<-tb_vax_AB
colnames(d01)<-c("samplemonth", "prop.vax", "data")
d01$prop.vax = d01$prop.vax/100
```

@@

```{r}
# Estimated vaccination rate: by anti-S + and anti-N neg
View(tb_vax_est)
d02<-select(tb_vax_est, samplemonth, prop.vax)
d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))

# Estimated vaccination rate: by anti-S only
View(tb_abbs)
d03<-select(tb_abbs, samplemonth, prop.abbs1)
colnames(d03)<-c("samplemonth", "prop.vax")
d03$data<-rep("S_pos", length(d03$samplemonth))

@@

# merge all
dta.vax<-rbind.data.frame(d01, d02, d03, d04)
rm(d01, d02, d03, d04)
dta.vax$prop.vax<-as.numeric(dta.vax$prop.vax)
```

@@

```{r}
p12<-dta.vax %>%
  ggplot(aes(x=samplemonth, y=prop.vax, group=data, colour=data)) +
  geom_point(aes(colour=data)) +
  geom_line(aes(colour=data))  +
  theme_bw() + 
  ylim(0,1)+
  ylab("Vaccinated%") +
  scale_colour_manual("", values = c("orange", "red", "red", "purple"),
                      labels = c("AB General Pop.", "APL Vax%", 
                                 "Anti-S(+)", "Anti-S(+) N(-)")
                      )
#  scale_linetype_manual("", values=c(3,3,1,1), labels = p09.lab) 
p12
```

#### 4.2.4 Apply (S+ N-)/N- for vax% estimation

First we calculate (S+ N-)/N- among CBS samples
```{r}
table(CBSs_AB$pos_roche, CBSs_AB$pos_roche_s, useNA = 'ifany')
# good data quality, only two NAs in roche_n and roche_s

# remove two NAs
CBSs_AB<-CBSs_AB[!(is.na(CBSs_AB$pos_roche) | is.na(CBSs_AB$pos_roche_s)), ]

# add vax_est
CBSs_AB$vax_est<-rep(0, length(CBSs_AB$pid))
CBSs_AB$vax_est[CBSs_AB$pos_roche==0 & CBSs_AB$pos_roche_s==1]<-1
table(CBSs_AB$vax_est)
```

Then we calculate the vax% using S(+) and (S+ & N-)/(N-)
```{r}
tbCBS_vax_est<-CBSs_AB %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_pos = sum(pos_roche, na.rm = T),
            s_pos = sum(pos_roche_s, na.rm = T),    
            n_vax= sum(vax_est, na.rm = T)
            )
tbCBS_vax_est$n_neg = tbCBS_vax_est$n - tbCBS_vax_est$n_pos
tbCBS_vax_est$prop.vax<-tbCBS_vax_est$n_vax/tbCBS_vax_est$n_neg

## add CBS data to merged data
d05<-select(tbCBS_vax_est, c(1,7))
colnames(d05)<-c("samplemonth", "prop.vax")
d05$data<-rep("CBS: S_pos_N_Neg", length(d05$samplemonth))

dta.vax<-rbind.data.frame(dta.vax, d05)
```

#### Now redo the plot
```{r}
# Add study and adjustment
dta.vax$study[dta.vax$data %in% c("AB_general")]<-"AB_general"
dta.vax$study[dta.vax$data %in% c("CBS: S_pos_N_Neg")]<-"CBS"
dta.vax$study[dta.vax$data %in% c("APL_vax", "S_pos", "S_pos_N_Neg")]<-"APL"

dta.vax$adj[dta.vax$data %in% c("AB_general", "APL_vax")]<-"raw"
dta.vax$adj[dta.vax$data %in% c("CBS: S_pos_N_Neg", "S_pos_N_Neg")]<-"S_pos_N_Neg"
dta.vax$adj[dta.vax$data %in% c("S_pos")]<-"S_pos"

p12.lab <- c("AB General Pop.", "APL Vax%", 
             "APL: Anti-S(+)", "APL: Anti-S(+) N(-)", "CBS: Anti-S(+) N(-)")
p12<-dta.vax %>%
  ggplot(aes(x=samplemonth, y=prop.vax, colour=interaction(study, adj),
                                        linetype=interaction(study, adj))) +
  geom_point() +
  geom_line()  +
  theme_bw() + 
  ylim(0,1)+
  ylab("Vaccinated%") +
  scale_colour_manual("", values = c("orange", "red", "red","purple", "blue"),
                      labels = p12.lab)+
  scale_linetype_manual("", values=c(1,1,3,3,3), labels = p12.lab) 
p12
```
#### Print p12
```{r}
png(file="../4_output/figs/Fig.S2 Vax.png", 
    width=2400, height=1440, res=300)
p12
dev.off()
```
